OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2018-10-31 phil_paper/jmp_qje.tex
send to dan

%\usepackage[dvipsnames]{xcolor}
%\usepackage{cite}
\onehalfspacing
\usepackage{natbib}  
\title{ Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila }
\author{William Violette}
\affil{Brown University\thanks{ I am grateful to Andrew Foster, Matthew Turner, Jesse Shapiro, Daniel Bj\"{o}rkegren, and Bryce Steinberg for guidance and encouragement.  For helpful conversations, I am also grateful to John Friedman, Edward Glaeser, Emily Oster, Brian Knight, and Lint Barrage.  In Manila, thank you to the staff of my water company partner, government regulators at MWSS, as well as the USAID Be-Secure project for access to data and excellent conversations.  This work was supported through the Watson Institute for International and Public Affairs. } \\ [Job Market Paper: \href{https://drive.google.com/file/d/1dM41qZLZY3EPtUqHbca6MIS-nT77hv_-/view?usp=sharing}{\textcolor{blue}{Link to Current Version}}]}

@@

	Public utilities subsidize fixed connection fees with high marginal prices to provide access especially for the poor, but this policy can have the opposite effect when many households share connections.  Using data on 1.5 million water connections in Manila, I establish that informal sharing networks provide \input{tables/shr_use_from_neighbor}\unskip\% 
	of total access.  I structurally estimate household water demand across three sources: purchasing directly from the provider, sharing with a neighbor's tap, or buying from a small-scale vendor.  The model predicts that low fixed fees and high marginal prices decrease access by weakening incentives to share water.  In contrast, the optimal pricing policy features high fixed fees and low marginal prices, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.  This policy improves welfare by up to \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income.

@@

Water sharing has important implications for the welfare impacts of common pricing policies.  Increasing marginal prices weaken incentives for households to extend water access to their neighbors through sharing (\cite{whittington1992possible}).  Fixed fee subsidies may have limited impacts on water access when sharing households split connection fees with their neighbors.  Since low-income households are more likely to use water from their neighbors, increasing marginal prices and low fixed fees may combine to act as regressive taxes in the presence of sharing.

@@

Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income.  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections to install and maintain.  Yet, since sharing involves costly redistribution of water, the optimal tariff subsidizes individual connections on the margin by pricing above marginal cost and setting the fixed fee below the cost of installing and maintaining each connection.

@@

In focusing on water sharing behavior, this approach abstracts away from several features that may also affect optimal water pricing.  First, I do not directly model environmental, health, or other externalities associated with piped water use.  Water shortages may lead policymakers to promote conservation through higher marginal prices while positive health externalities may motivate further subsidizing piped water (\cite{timmins2002measuring}, \cite{galiani2005water}).\footnote{\cite{bennett2012does} shows that piped water can also lead to negative externalities through worse sanitation, possibly motivating higher prices for piped water in some cases.}  Second, I assume that government regulator is able to fully observe production costs for the water provider, eliminating scope for prices to correct for possible principal-agent frictions as characterized by \cite{laffont1993theory}.  Finally, the counterfactual exercises borrow the public-private partnership regulatory framework in Manila by assuming that revenue must fully cover production costs.  Additional data on the costs of raising public funds and risks of regulatory capture would allow for a more flexible analysis taking into account the possibility of public funding for water (\cite{laffont2005regulation}).

@@

The next section describes the context of water provision in Manila.  Section~\ref{section:data} describes the data.  Section~\ref{section:patternsofwateruse} presents stylized facts about water usage in Manila.  Section~\ref{section:modelofhouseholdwateruse} introduces a model of household water usage and source choices.  The estimation is divided into three steps in Section~\ref{section:specificationandestimation}.  Section~\ref{section:preferenceestimation} uses monthly quantity choices and price variation to estimate preferences.  Section~\ref{section:hasslecostestimation} analyzes a quasi-experiment to recover an estimate of the marginal costs of sharing with a neighbor.  Given these estimates, Section~\ref{section:fixedcostestimation} recovers the relative fixed costs of owning a connection or using from a vendor.  Section~\ref{section:counterfactualexercises} uses these estimates to examine counterfactual pricing policies.  Section~\ref{section:conclusion} concludes.

@@

Supplemental survey data from a subset of Manila include income measures that are used to approximate household income using the full set of demographic measures and their interactions.

@@

In the first step, each household makes a one-time decision about their water source between 1) purchasing a connection, 2) using a neighbor's connection, or 3) buying from a local water vendor.  These decisions result in a set of water connections for each neighborhood ranging from an empty set (if all households use from a vendor) to the whole neighborhood (if each household uses an individual connection).  Let each connection, $C$, serve $S$ households that have decided to share the connection.  For ease of notation, let the subscript $C$ indicate connection-level attributes, summing over all households using from connection $C$.

This model does not allow for households to choose different water sources over time, which is consistent with large fixed costs involved in initially purchasing a connection or establishing a sharing relationship.  Section~\ref{section:hasslemodel} extends the model to allow for cases where households exogenously lose access to their current water source and must substitute toward another nearby source.

In the second step, each household then chooses how much water to use each month from their source.  Since water sources provide utility through monthly consumption, the following section describes the decision process in reverse order, first characterizing monthly water consumption choices (taking the initial source choice as given) and second, describing the initial water source choices.

@@

As a simple way to capture this coordination task, let each household $i$ set a goal for water use each month, $w_i$, knowing that unobserved consumption shocks, $\epsilon_i$, may increase or decrease their realized amount of usage.  Consumption shocks, $\epsilon_i$, are assumed to be unobserved to the household and independent across time, bringing the total realized water consumption each month to, $W_i=w_i+\epsilon_i$.  Household utility over total water, $W_i$ and a numeraire good, $x_i$ is assumed to take the following shape:
\begin{align}\label{equation:umaxfirst}
\begin{split}
U(x_i,W_i) &=  x_i + (1-H_i) W_i - \frac{1}{2\alpha_i } ( W_i - \Gamma_i + \alpha_i )^2  \\
W_i &= w_i + \epsilon_i \\
\epsilon_i &\sim \mathcal{N}(0,\Sigma_{\epsilon})
\end{split}
\end{align}
Following from quasi-linear preferences, this quadratic form for utility eliminates income effects and includes a natural satiation point for water consumption, capturing the tendency of households to target specific consumption levels despite low or zero marginal prices as in Figure~\ref{figure:consumptionhistogram}.\footnote{This utility form builds on \cite{einav2013selection} who use a similar form to characterize extensive and intensive choices of health insurance.}   Household preferences for water are characterized by $\Gamma_i$, which can be interpreted as a fixed preference for water each month and $\alpha_i$, which captures the extent to which households are price sensitive.  The exogenous shock to consumption, $\epsilon_i$, is assumed to be drawn from a multivariate normal distribution, $\Sigma_{\epsilon}$ with zero mean and variance terms, $\sigma_{\epsilon}^2$, allowing for correlation with neighboring households $i$ and $j$ through term ${\rho}^{\epsilon}_{i,j}$.  Although households cannot anticipate the exact realization of $\epsilon_i$, they are assumed to understand the distribution of possible consumption shocks, allowing them to maximize expected utility.

@@

To account for idiosyncratic preference shocks for water each month, let household's fixed preference, $\Gamma_{i}$, be decomposed into a static, household-specific component, $\gamma_i$, and a normally distributed, idiosyncratic component, $\eta_{i}$, which is assumed to follow a multivariate normal distribution among sharing partners:
\begin{align*}
\Gamma_{i} &= \gamma_i + \eta_{i} \\
\eta_{i} &\sim \mathcal{N} (\textbf{0},\Sigma_{\eta}) 
\end{align*}
This preference shock is also assumed to be independent of the consumption shocks, $\epsilon_C$.  Aggregating these preferences shocks over all sharing households using connection, $C$, produces a joint shock, $\eta_C$, according to the convolution of these random normal variables with a variance equal to $\sum_{i=1}^{S} \sigma_{\eta}^2 + \sum_{i=1}^{S} \sum_{j>i}^{S} \rho^{\eta}_{i,j} \sigma_{\eta}^2$.

@@

The household problem is additionally complicated by the presence of non-linear price schedules for water.  Figure~\ref{figure:tariff} provides the increasing tariff in Manila.  The non-linear price schedule produces kinks in household budget constraints leading to non-linear demand functions.  Let $P(W_C) = \{ p^{1},...,p^{K},\overline{W}^{0},...,\overline{W}^{K}\}$ describe tariff schedule as a function of total consumption for the water connection, $W_C = \sum_{i=1}^{S} W_i$.  The tariff includes $K$ segments where segment $k$ corresponds to price $p^{k}$ for water consumption between the consumption thresholds (or kink points), $\overline{W}^{k-1}$ and $\overline{W}^{k}$:\footnote{In this case in Manila, $\overline{W}^{0}$ is zero while $\overline{W}^{K}$ is infinity.}
\begin{align*}
P(W_C)=&  
\begin{cases}
      p^{1} &\text{ if } W_C \in [\overline{W}^{0},\overline{W}^{1}], \\
      p^{2} &\text{ if } W_C \in [\overline{W}^{1},\overline{W}^{2}], \\
      ...   &\text{ ...}, \\
      p^{K} &\text{ if } W_C \in[\overline{W}^{K-1},\overline{W}^{K}]
    \end{cases}
\end{align*}
Given prices and household utility, the final ingredient for characterizing consumption choices is the budget constraint.  The budget constraint includes monthly household income, $Y_i$, and monthly fixed costs of using the water connection, $F_i$, which are assumed to capture three components: (1) monthly service fees, (2) initial connection fees amortized in monthly terms for the duration of the connection, and (3) unobserved, fixed hassle costs associated with setting up a connection, which are also amortized into monthly terms.  Data does not track household water source choices over time in a comprehensive way, which limits the extent to which the model can distinguish between one-time connection costs and monthly service/maintenance costs.  This limitation means that the model cannot address the optimal timing of fixed payments or capture potential heterogeneity in monthly fixed costs.

Sharing households must negotiate over how to pay the monthly fixed costs (including the fixed fee paid to the firm and any unobserved costs of using a connection).  The model assumes that households commit to dividing any fixed costs at the start of the sharing arrangement, allocating $F_i$ to each household.  Under quasi-linear utility, any fixed transfers are exogenous to monthly consumption decisions and Section~\ref{section:watersourcechoice} later discusses how fixed costs are divided.

Water sharing relationships impact monthly consumption choices through the budget constraint.\footnote{While the formation of these relationships is discussed in more detail in Section~\ref{section:watersourcechoice}, this section takes these sharing relationships as exogenously given.}  The model assumes that the monthly water bill is split evenly between all sharing households.  Descriptive evidence suggests that challenges observing and coordinating usage within members of a household also extend across households.  Among owners of the water connection responding to the mobile survey, \input{tables/bill_split}\unskip\% report splitting the water bill evenly or alternating payment each month, \input{tables/bill_fixed}\unskip\% charge sharing households a fixed amount each month regardless of their consumption, and \input{tables/bill_usage}\unskip\% charge other users a price according to their usage.\footnote{The remaining households report ``I don't know'' or ``other.''}  Evenly splitting the bill, alternating payment, or charging a fixed amount provide easy ways to extract compensation without requiring specific information on the monthly consumption of each sharing household.  Yet, these methods also introduce free-rider incentives since households do not have to fully pay for their contributions to the total water bill.  Since data limitations prevent identifying the specific type of contract associated with each shared connection, the model assumes that all households evenly split the bill and explores the theoretical and empirical implications of alternate models in Appendix~\ref{appendix:fixedprice}.  The model captures bill splitting by dividing the total monthly payment for a water connection across the $S$ sharing households.

@@

Normalizing the price of the numeraire good, $x$, to one, the budget constraint for household, $i$, using a connection with households, $S$, takes the following form given that total consumption for their connection, $W_C$, falls between thresholds, $\overline{W}^{k-1}$ and $\overline{W}^{k}$:
%\begin{align*}
%Y_i - F_i &=  
%\begin{cases}
%      x_i + H_{i} W_i + \frac{1}{S} p^{1} W_C   &\text{ if }   W_C \in [0,\overline{W}^{1}], \\
%      x_i +  H_{i} W_i +  \frac{1}{S} \big [p^{k} W_C -  E_{k} \big]     &\text{ if } W_C \in [\overline{W}^{k-1},\overline{W}^{k}], \\
%      ...   &\text{ ... }      
%\end{cases} \\
%E_{k} &= \big  [\sum_{j=1}^{k-1}  (p^{k} - p^{j})(\overline{W}^{j}-\overline{W}^{j-1})  \big ]
%\end{align*}
\begin{align}
\label{equation:budgetconstraint}
\begin{split}
(Y_i - F_i) &= x_i +   \dfrac{  B_{C,k}  }{S_C}    \hspace{.3cm} \text{ if } \hspace{.1cm} W_C \in [\overline{W}^{k-1},\overline{W}^{k}] \\
\text{Where: }& \\
B_{C,k}  &= p^{k} (W_{C}^{*} - \overline{W}^{k-1}) +\sum_{j=1}^{k-1}  p^{j}(\overline{W}^{j}-\overline{W}^{j-1}) 
\end{split}
\end{align}

The full budget constraint is a piece-wise expansion of equation (\ref{equation:budgetconstraint}) over all $K$ price thresholds.  $B_{C,k}$ represents the total water bill for connection $C$ given the tariff schedule.  This term is calculated by subtracting the current price that the household faces, $p^{k}$, from the previous prices and multiplying by their exposure given by, $(\overline{W}^{j}-\overline{W}^{j-1})$.
%\footnote{In the case where a household uses a connection individually, this expression simplifies by setting the index, $i$, and the total using households, $S$, both equal to one.  In this scenario, the household does not face any hassle costs, $H_i$, and pays the total bill from its corresponding consumption.}

@@

When households share a water connection, the consumption choice for each household becomes a strategic decision.  Since prices are non-linear, the total consumption level for each connection, $W_C$, determines the marginal price faced by all households.  For example, with an increasing block tariff like Figure~\ref{figure:tariff}, one household's choice to increase consumption may produce a higher marginal price for other sharing households.

Substituting the numeraire good, $x_i$ for the budget constraint in household utility from equation (\ref{equation:umaxfirst}) simplifies the household's problem to a single choice variable --- targeted consumption, $w_i$.  The household's expected utility problem over consumption shocks $\epsilon_C$ then takes the following form:

%\scalebox{.88}{\parbox{1\linewidth}{
%\begin{align*}
%E[U(x_i,w_i)] = &  (Y_i - F_i)  - \frac{1}{S} \sum_{k=1}^{K} \int_{\overline{W}^{k-1}-W_C}^{\overline{W}^{k}-W_C}\big [p^{k} W_C -  E_{k} \big] d \epsilon_C + \int_{-\infty}^{\infty} W_i - H_{i} W_i - \frac{1}{2\alpha_i} ( W_i - \Gamma_i + \alpha_i )^2 d \epsilon_i  \\  
 %         = &  (Y_i - F_i)  - \frac{1}{S} \sum_{k=1}^{K} \int_{\overline{W}^{k-1}-W_C}^{\overline{W}^{k}-W_C}\big [p^{k} W_C d -  E_{k} \big] d \epsilon_C+ w_i - H_{i} w_i - \frac{1}{2\alpha_i} ( w_i - \Gamma_i + \alpha_i )^2 + \sigma_{\epsilon}^{2}  
%\end{align*}
%}} \\

% E[U(w_i)] = 
\begin{align*}
max_{w_i} \sum_{k=1}^{K} \int_{\overline{W}^{k-1}-{w_C}}^{\overline{W}^{k}-{w_C}} (Y_i - F_i) -  \dfrac{ B_{C,k} }{S_C}     + (1-p_s) W_i   - \frac{1}{2\alpha } ( W_i - \Gamma_i + \alpha_i )^2  d \epsilon_C 
\end{align*}

>>>>

\usepackage[notablist,nofiglist]{endfloat}
%\renewcommand{\efloatseparator}{\mbox{}}

\renewcommand{\thetable}{\Roman{table}}
\renewcommand{\thefigure}{\Roman{figure}}

\renewcommand{\thesection}{\Roman{section}} 
\renewcommand{\thesubsection}{\thesection.\Alph{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\roman{subsubsection}}

%\usepackage[dvipsnames]{xcolor}
%\usepackage{cite}
\onehalfspacing
\usepackage{natbib}  
%\title{ Optimal Pricing and Informal Sharing: \\ Evidence from Piped Water in Manila }
\title{ OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA}

\author{William Violette\thanks{Contact: Federal Trade Commission, Washington, DC: william.j.violette@gmail.com \\ I am grateful to Andrew Foster, Matthew Turner, Jesse Shapiro, Daniel Bj\"{o}rkegren, and Bryce Steinberg for guidance and encouragement.  For helpful conversations, I am also grateful to John Friedman, Edward Glaeser, Emily Oster, Brian Knight, and Lint Barrage.  In Manila, thank you to the staff of my water company partner, government regulators at MWSS, as well as the USAID Be-Secure project for access to data and excellent conversations.  This work was supported through the Watson Institute for International and Public Affairs.  Any opinions and conclusions expressed herein are those of the author and do not necessarily represent the views of the the Federal Trade Commission or its Commissioners.}
 }
%\\ 
%[Job Market Paper: \href{https://drive.google.com/file/d/1dM41qZLZY3EPtUqHbca6MIS-nT77hv_-/view?usp=sharing}{\textcolor{blue}{Link to Current Version}}]}

@@

	Public utilities subsidize fixed connection fees with high marginal prices to provide access especially for the poor, but this policy can have the opposite effect when many households share connections.  Using data on 1.5 million water connections in Manila, I establish that informal sharing networks provide \input{tables/shr_use_from_neighbor}\unskip\% 
	of total access.  I structurally estimate household water demand across three sources: purchasing directly from the provider, sharing with a neighbor's tap, or buying from a small-scale vendor.  The model predicts that low fixed fees and high marginal prices decrease access by weakening incentives to share water.  In contrast, the optimal pricing policy features a high fixed fee and low marginal price, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.  This policy improves welfare by up to \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income.

@@

Water sharing has important implications for the welfare impacts of common pricing policies.  Increasing marginal prices weaken incentives for households to extend water access to their neighbors through sharing (\cite{whittington1992possible}).  Fixed fee subsidies may have limited impacts on water access when sharing households already split connection fees with their neighbors.  Since low-income households are more likely to use water from their neighbors, increasing marginal prices and low fixed fees may combine to act as regressive taxes in the presence of sharing.

@@

Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income.  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections to install and maintain.  
%Yet, since sharing involves costly redistribution of water, the optimal tariff subsidizes individual connections on the margin by pricing above marginal cost and setting the fixed fee below the cost of installing and maintaining each connection.

@@

In focusing on water sharing behavior, this approach abstracts away from several features that may also affect optimal water pricing.  First, I do not directly model environmental, health, or other externalities associated with piped water use.  Water shortages may lead policymakers to promote conservation through higher marginal prices while positive health externalities may motivate further subsidizing piped water (\cite{timmins2002measuring}, \cite{galiani2005water}).\footnote{\cite{bennett2012does} shows that piped water can also lead to negative externalities through worse sanitation, possibly motivating higher prices for piped water in some cases.}  Second, I assume that government regulator is able to fully observe production costs for the water provider, eliminating scope for prices to correct for possible principal-agent frictions as characterized by \cite{laffont1993theory}.  Finally, the counterfactual exercises simulate the public-private partnership regulatory framework in Manila by assuming that revenue must fully cover production costs.  Additional data on the costs of raising public funds and risks of regulatory capture would allow for a more flexible analysis taking into account the possibility of using public funding for water infrastructure (\cite{laffont2005regulation}).

@@

This paper proceeds by first describing the context of water provision in Manila.  Section~\ref{section:data} describes the data.  Section~\ref{section:patternsofwateruse} presents stylized facts about water usage in Manila.  Section~\ref{section:modelofhouseholdwateruse} introduces a model of household water usage and source choices.  The estimation is divided into three steps in Section~\ref{section:specificationandestimation}.  Section~\ref{section:preferenceestimation} uses monthly quantity choices and price variation to estimate preferences.  Section~\ref{section:hasslecostestimation} analyzes a quasi-experiment to recover an estimate of the marginal costs of sharing with a neighbor.  Given these estimates, Section~\ref{section:fixedcostestimation} recovers the relative fixed costs of owning a connection or using from a vendor.  Section~\ref{section:counterfactualexercises} uses these estimates to examine counterfactual pricing policies.  Section~\ref{section:conclusion} concludes.

@@

With information on water source and monthly expenditures as well as household income, the 2011 Community Based Monitoring System data cover over 70\% of households in Pasay City --- a large area with a population of 500,000 in downtown metro Manila.  These data are used to impute income for households outside of the sample according to the following demographic measures: age of household head and indicators for low-skilled employment of household head, residing in a duplex, residing in a house, as well as all possible household sizes and number of employed household members.  The imputation also includes all possible interactions between these measures.

%Supplemental survey data from the Community Based Monitoring Survey in Pasay City --- a large city in the center of Metro Manila --- include household income measures for  that are used to approximate household income using the full set of demographic measures and their interactions.

@@

In the first step, each household makes a one-time decision about their water source between 1) purchasing a connection, 2) using a neighbor's connection, or 3) buying from a local water vendor.  These decisions result in a set of water connections for each neighborhood ranging from an empty set (if all households use from a vendor) to the whole neighborhood (if each household uses an individual connection).  Let each connection, $C$, serve $S$ households that have decided to share the connection.  For ease of notation, let the subscript $C$ indicate connection-level attributes, summing over all households using from connection $C$.  This model does not allow for households to choose different water sources over time, which is consistent with large fixed costs involved in initially purchasing a connection or establishing a sharing relationship.  Section~\ref{section:hasslemodel} extends the model to allow for cases where households may exogenously lose access to their current water source and must substitute toward another nearby source.

In the second step, each household then chooses how much water to use each month from their source.  Since water sources provide utility through monthly consumption, it is useful to describe the decision process in reverse order, first characterizing monthly water consumption choices (taking the initial source choice as given) and second, characterizing the initial water source choices.

@@

As a simple way to capture this coordination task, let each household $i$ choose a goal for water use each month, $w_i$, knowing that unobserved consumption shocks, $\epsilon_i$, may increase or decrease their realized amount of usage.  Consumption shocks, $\epsilon_i$, are assumed to be unobserved to the household and independent across time, bringing the total realized water consumption each month to, $W_i=w_i+\epsilon_i$.  Household utility over total water, $W_i$ and a numeraire good, $x_i$ is assumed to take the following shape:
\begin{align}\label{equation:umaxfirst}
\begin{split}
U(x_i,W_i) &=  x_i + (1-H_i) W_i - \frac{1}{2\alpha_i } ( W_i - \Gamma_i + \alpha_i )^2  \\
W_i &= w_i + \epsilon_i \\
\epsilon_i &\sim \mathcal{N}(0,\Sigma_{\epsilon})
\end{split}
\end{align}
Following from quasi-linear preferences, this quadratic form for utility eliminates income effects and includes a natural satiation point for water consumption, capturing the tendency of households to target specific consumption levels despite low or zero marginal prices as shown in Figure~\ref{figure:consumptionhistogram}.\footnote{This utility form builds on work by \cite{einav2013selection} who use a similar form to characterize extensive and intensive choices of health insurance.}   Household preferences for water are characterized by $\Gamma_i$, which can be interpreted as a fixed preference for water each month and $\alpha_i$, which captures the extent to which households are sensitive to price.  The exogenous shock to consumption, $\epsilon_i$, is assumed to be drawn from a multivariate normal distribution, $\Sigma_{\epsilon}$ with zero mean and variance terms, $\sigma_{\epsilon}^2$, allowing for correlation with neighboring households $i$ and $j$ through term ${\rho}^{\epsilon}_{i,j}$.  Although households cannot anticipate the exact realization of $\epsilon_i$, they are assumed to understand the distribution of possible consumption shocks, allowing them to maximize expected utility.

@@

To account for idiosyncratic preference shocks for water each month, let household's fixed preference, $\Gamma_{i}$, be decomposed into a static, household-specific component, $\gamma_i$, and an idiosyncratic component, $\eta_{i}$, which is assumed to follow a multivariate normal distribution among sharing partners:
\begin{align*}
\Gamma_{i} &= \gamma_i + \eta_{i} \\
\eta_{i} &\sim \mathcal{N} (\textbf{0},\Sigma_{\eta}) 
\end{align*}
This preference shock is also assumed to be independent of the consumption shocks, $\epsilon_C$.  Aggregating these preferences shocks over all households sharing connection, $C$, produces a joint shock, $\eta_C$, according to the convolution of these random normal variables with a variance equal to $\sum_{i=1}^{S} \sigma_{\eta}^2 + \sum_{i=1}^{S} \sum_{j>i}^{S} \rho^{\eta}_{i,j} \sigma_{\eta}^2$.

@@

The household problem is additionally complicated by the presence of an increasing price schedule for water; as a household uses more water in one month, they face higher marginal prices.  Figure~\ref{figure:tariff} provides the increasing tariff in Manila.  This price schedule produces kinks in household budget constraints leading to non-linear demand functions.  Let $P(W_C) = \{ p^{1},...,p^{K},\overline{W}^{0},...,\overline{W}^{K}\}$ describe the tariff schedule as a function of total consumption for the water connection, $W_C = \sum_{i=1}^{S} W_i$.  The tariff includes $K$ segments where segment $k$ corresponds to price $p^{k}$ for water consumption between the consumption thresholds (or kink points), $\overline{W}^{k-1}$ and $\overline{W}^{k}$:\footnote{In this case in Manila, $\overline{W}^{0}$ is zero while $\overline{W}^{K}$ is infinity.}
\begin{align*}
P(W_C)=&  
\begin{cases}
      p^{1} &\text{ if } W_C \in [\overline{W}^{0},\overline{W}^{1}], \\
      p^{2} &\text{ if } W_C \in [\overline{W}^{1},\overline{W}^{2}], \\
      ...   &\text{ ...}, \\
      p^{K} &\text{ if } W_C \in[\overline{W}^{K-1},\overline{W}^{K}]
    \end{cases}
\end{align*}
Given prices and household utility, the final ingredient for characterizing consumption choices is the budget constraint.  The budget constraint includes monthly household income, $Y_i$, and monthly fixed costs of using the water connection, $F_i$, which are assumed to capture three components: (1) monthly service fees, (2) initial connection fees amortized into monthly terms for the duration of the connection, and (3) unobserved, fixed hassle costs associated with setting up a connection, which are also amortized into monthly terms.  Data does not track household water source choices over time in a comprehensive way, which limits the extent to which the model can distinguish between one-time connection costs and monthly service/maintenance costs.  This limitation means that the model cannot address the optimal timing of fixed payments or capture potential heterogeneity in monthly fixed costs.

Sharing households must negotiate over how to pay the monthly fixed costs (including the fixed fee paid to the firm and any unobserved costs of using a connection).  The model assumes that households commit to dividing any fixed costs at the start of the sharing arrangement, allocating $F_i$ to each household.  Under quasi-linear utility, any fixed transfers are exogenous to monthly consumption decisions and Section~\ref{section:watersourcechoice} later discusses how fixed costs are divided between sharing households.

Water sharing relationships impact monthly consumption choices through the budget constraint.\footnote{While the formation of these relationships is discussed in more detail in Section~\ref{section:watersourcechoice}, this section takes these sharing relationships as exogenously given.}  The model assumes that the monthly water bill is split evenly between all sharing households.  Descriptive evidence suggests that challenges observing and coordinating usage within members of a household also extend across households.  Among owners of the water connection responding to the mobile survey, \input{tables/bill_split}\unskip\% report splitting the water bill evenly or alternating payment each month, \input{tables/bill_fixed}\unskip\% charge sharing households a fixed amount each month regardless of their consumption, and \input{tables/bill_usage}\unskip\% charge other users a price according to their usage.\footnote{The remaining households report ``I don't know'' or ``other.''}  Evenly splitting the bill, alternating payment, or charging a fixed amount provide easy ways to extract compensation without requiring specific information on the monthly consumption of each sharing household.  Yet, these methods also introduce free-rider incentives since households do not have to fully pay for their contributions to the total water bill.  Since data limitations prevent identifying the specific type of contract associated with each shared connection, the model assumes that all households evenly split the bill and explores the theoretical and empirical implications of alternate models in Appendix~\ref{appendix:fixedprice}.  The model captures bill splitting by evenly dividing the total monthly payment for a water connection across the $S$ sharing households.

@@

Normalizing the price of the numeraire good, $x$, to one, the budget constraint for household, $i$, using a connection with households, $S$, takes the following form given that total consumption for their connection, $W_C$, falls between thresholds, $\overline{W}^{k-1}$ and $\overline{W}^{k}$:
%\begin{align*}
%Y_i - F_i &=  
%\begin{cases}
%      x_i + H_{i} W_i + \frac{1}{S} p^{1} W_C   &\text{ if }   W_C \in [0,\overline{W}^{1}], \\
%      x_i +  H_{i} W_i +  \frac{1}{S} \big [p^{k} W_C -  E_{k} \big]     &\text{ if } W_C \in [\overline{W}^{k-1},\overline{W}^{k}], \\
%      ...   &\text{ ... }      
%\end{cases} \\
%E_{k} &= \big  [\sum_{j=1}^{k-1}  (p^{k} - p^{j})(\overline{W}^{j}-\overline{W}^{j-1})  \big ]
%\end{align*}
\begin{align}
\label{equation:budgetconstraint}
\begin{split}
(Y_i - F_i) &= x_i +   \dfrac{  B_{C,k}  }{S_C}    \hspace{.3cm} \text{ if } \hspace{.1cm} W_C \in [\overline{W}^{k-1},\overline{W}^{k}] \\
\text{Where: }& \\
B_{C,k}  &= p^{k} (W_{C}^{*} - \overline{W}^{k-1}) +\sum_{j=1}^{k-1}  p^{j}(\overline{W}^{j}-\overline{W}^{j-1}) 
\end{split}
\end{align}
The full budget constraint is a piece-wise expansion of equation (\ref{equation:budgetconstraint}) over all $K$ price thresholds.  $B_{C,k}$ represents the total water bill for connection $C$ given the tariff schedule.  This term is calculated by multiplying each price by the volume of consumption exposed to that price.
%\footnote{In the case where a household uses a connection individually, this expression simplifies by setting the index, $i$, and the total using households, $S$, both equal to one.  In this scenario, the household does not face any hassle costs, $H_i$, and pays the total bill from its corresponding consumption.}

@@

When households share a water connection, the consumption choice for each household becomes a strategic decision.  Since prices are non-linear, the total consumption level for each connection, $W_C$, determines the marginal price faced by all households.  For example, with an increasing tariff like Figure~\ref{figure:tariff}, one household's choice to increase consumption may produce a higher marginal price for other sharing households.

Substituting the budget constraint (equation (\ref{equation:budgetconstraint})) into the household utility (equation (\ref{equation:umaxfirst})) for the numeraire good $x_i$ simplifies the household's maximization problem to a single choice variable --- targeted consumption, $w_i$.  The household's expected utility problem over consumption shocks $\epsilon_C$ then takes the following form:
%\scalebox{.88}{\parbox{1\linewidth}{
%\begin{align*}
%E[U(x_i,w_i)] = &  (Y_i - F_i)  - \frac{1}{S} \sum_{k=1}^{K} \int_{\overline{W}^{k-1}-W_C}^{\overline{W}^{k}-W_C}\big [p^{k} W_C -  E_{k} \big] d \epsilon_C + \int_{-\infty}^{\infty} W_i - H_{i} W_i - \frac{1}{2\alpha_i} ( W_i - \Gamma_i + \alpha_i )^2 d \epsilon_i  \\  
 %         = &  (Y_i - F_i)  - \frac{1}{S} \sum_{k=1}^{K} \int_{\overline{W}^{k-1}-W_C}^{\overline{W}^{k}-W_C}\big [p^{k} W_C d -  E_{k} \big] d \epsilon_C+ w_i - H_{i} w_i - \frac{1}{2\alpha_i} ( w_i - \Gamma_i + \alpha_i )^2 + \sigma_{\epsilon}^{2}  
%\end{align*}
%}} \\
% E[U(w_i)] = 
\begin{align*}
max_{w_i} &\sum_{k=1}^{K} \int_{\overline{W}^{k-1}-{w_C}}^{\overline{W}^{k}-{w_C}} (Y_i - F_i) -  \dfrac{ B_{C,k} }{S_C}     + (1-p_s) W_i   - \frac{1}{2\alpha } ( W_i - \Gamma_i + \alpha_i )^2  d \epsilon_C  
\end{align*}
%B_{C,k}  &= p^{k} (W_{C}^{*} - \overline{W}^{k-1}) +\sum_{j=1}^{k-1}  p^{j}(\overline{W}^{j}-\overline{W}^{j-1}) 
Where $\epsilon_C$ is the convolution of consumption shocks across sharing households distributed as follows:
\begin{align*}
          \epsilon_C &\sim \mathcal{N}(0,\sigma_{\epsilon_C}^2 = \sum_{i=1}^{S} \sigma_{\epsilon}^2 + \sum_{i=1}^{S} \sum_{j>i}^{S} \sigma_{\epsilon}^2 \rho^{\epsilon}_{i,j} )
\end{align*}
To solve this system, this approach assumes that households follow a perfect information, pure-strategy Nash-Equilibrium in choosing consumption.  Using this structure, households compute their best response consumption functions to arrive at a symmetric equilibrium taking consumption for their sharing partners as given.  
% % % % % % % % % % % % % % % % \footnote{  APPENDIX PROVES THIS;  NEED TO GO DEEPER PLEASE!!! SHOW EXPLICITLY }  % % % % % % % % % % % % % % % % % % %
%Because of the quadratic shape of the utility function, the exogenous shock to consumption, $\epsilon_s$, only affects the optimal consumption choice to the extent that it pushes the household across price thresholds.  Therefore, the solution to equation (\ref{equation:umaxfirst}) is isomorphic to removing the $\epsilon_s$ terms from the utility function and only allowing these shocks to enter through the budget constraint.  This simplification ensures that the $\epsilon_s$ terms enter linearly into the optimization problem through the budget constraint.
